Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-23 main_new.tex
Updates from Jo Eidsvik

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task for
  modern marine management and science faced with limited resources to
  understand highly dynamic processes in the water-column. The
  combination of statistics and autonomous
  robotics provides opportunities for studying aspects related to
  experimental designs for effective characterization of oceanographic
  phenomena. In this work we develop methods for efficient spatial
  sampling applied to the mapping of dynamic coastal processes by
  providing informative descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the excursion set of multivariate random
  fields, and derive analytical expressions for the expected Bernoulli
  variance reduction in the excursion set of spatial temperature and
  salinity variables. We show how these can be applicable in static
  designs and sequential designs, building on myopic and look-ahead
  sampling strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches. Results from field
  deployments with an autonomous underwater vehicle are also presented,
  for a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

\begin{itemize}
\item Decisions on where to sample happen sequentially, i.e.  how
  recently acquired information can be assimilated and used to improve
  statistical models and reduce uncertainty associated with where and
  when to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there lies a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}, this is
  usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the optimal design) and computability (arriving at a
  solution in reasonable time), which has to be considered in
  practice.
\end{itemize}{}

@@

In this work, we build on excursion sets (ES) and excursion
probabilities (EP) for spatial design of experiments. The goal is to
design sampling strategies for reducing the spatially integrated
Bernoulli variance (IBV) in the ES indicators.  The main contributions
involve: i) extending ES-based uncertainty criteria to multivariate
GPs, allowing the evaluation to include multiple environmental
parameters such as temperature and salinity, ii) developing methods
for informative environmental sensing and autonomous data collection,
and iii) demonstrating the approach in a real world scenario of
sampling a river plume using an AUV.

@@

AUVs are used to gather in-situ spatio-temporal data from a targeted
domain while carrying a range of scientific payloads to survey the
water column.  Typically, they operate at $\sim 1-3$ m/s, can reach
depths from 100--6000m with an in-water staying capacity depending on
survey speed, payload sensors and mission design. By deploying AUVs
one can fill gaps in data gathering and augment information obtained
from either ocean models, remote-sensing data, or fixed-location
sensors on buoys. Traditional operations and surveys with AUVs are
usually limited to observations along fixed transects on a spatial
grid, pre-programmed by the human operator.  By using on-board
algorithms to continuously evaluate, update, and refine future
sampling locations, one can make information gathering adaptive and
contextual to the survey requirements
\citep{das11b,fossuminformation,fossum18b}. The space of sampling
opportunities is still limited to the spatial grid, but now there is
much more flexibility because of adaptation. For instance, an AUV
could reconstruct or modify a survey line based on what temperatures
it is measuring by using embedded methods in decision-theoretic
planning \citep{py10,Rajan12,Rajan12b}.

@@

Our objective is to characterize a river plume, focusing on spatial
separation of cold freshwater from the warmer saline waters of the
fjord. Excursion sets (ES) and excursion probabilities (EP) are useful
starting points to measure the ability to classify water masses. The
goal of experimental sampling is to improve this characterization, as
quantified via changes in the EPs.

We consider a lateral domain, near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes;
$\xi_a(\bx)$ denotes the temperature (in $^o C$) and $\xi_b(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. \kc{might be
  clearer if you used 't' for temp and 's' for salinity. 'a' and 'b'
  are somewhat opaque especially further in the text. Also there
  should be a statement about why we're considering only 2 variables
  and the impact to considering (algebraically) more in the following
  formulations. This will surely be asked by any reviewer.}.  We let $t_a$ be a
threshold for temperature and $t_b$ a threshold for salinity, the
bivariate ES is then defined by:

\begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_a(\bx) \leq t_a,\xi_b(\bx) \leq t_b\}.
\end{equation}
The excursions could also be defined as larger than a threshold, or
within a boundary of limits for the two variables; calculations and
interpretations will be similar in these cases.

The ES holds random indicator variables, while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$, it is easy to classify water masses; it is
challenging if the probabilities are close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}
The measure is dominated by locations with EPs close to $0.5$.
\kc{This sentence is a bit strange; just about, you state that EPs
  close to 0.5 can be problematic since they do not categorize the
  water masses well. And here you state that there is a strong
  dominance of the average. So now you should say how the biasing can
  change to make ``V'' more effective. Otherwise this sentence is
  dangling.}

If prior knowledge provides a strong indication of cold freshwater
(for instance) in a part of the domain, it is easy to classify, and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification skill. We denote temperature measurements by
$y_{a}(\bx_i)$ and salinity measurements by $y_{b}(\bx_i)$, at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements, denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by). 
\end{equation}

Our goal is to construct designs for data gathering, that are expected
to improve the mapping and separation of water masses. When designing
AUV sampling designs, it is therefore natural to evaluate the expected
reduction in the IBV, as provided by the survey data. In this way, we
take the expectation with respect to the planned measurements $\by$;
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey. A criterion for the selection is:
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}), and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria in the situation with multivariate GPs.

@@

Calculations of the EPs and the IBV require a model specification.
Here, we use Gaussian modeling assumptions which allow closed form
evaluation of the IBVs, as derived in Section \ref{sec:sur}.

@@

We represent the bivariate random fields of temperature and salinity
with a Gaussian model. At location $\bx \in \mathcal{M}$, the
bivariate distribution is:
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_a(\bx) \\
    \xi_b(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{a}^2(\bx) & \sigma_{a}(\bx) \sigma_b(\bx) \gamma(\bx)  \\
\sigma_{a}(\bx) \sigma_b(\bx) \gamma(\bx)  & \sigma_{b}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (Normal) distributed variables
specified by the mean vector and covariance matrix. In the application
below, the model parameters are specified from preliminary data: The
mean varies with location, capturing the variability near the river
mouth, while the variance parameters and correlation between salinity
and temperature are assumed to be constant for all locations. \kc{?
  Isn't that too strong an assumption? T/S tend to be correlated,
  however for uniform bodies of water; here we're dealing with coastal
  ecosystems where the two are not necessarily so.}

For the spatial dependence among variables, 
we assume a separable correlation function with the same decay for salinity and temperature,
which is not unrealistic for a river plume \kc{why so?}. We have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
for $i,j \in {a,b}$ and $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$
otherwise. Moreover, we assume a stationary isotropic property
\kc{This is probably related to the assumption as above where T/S are
  assumed to be constant in all locations? But why?} where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via a Euclidean distance measure $h=\sqrt{|\bx-\bx'|^2}$. With
data and prior knowledge one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however that is left for
future work.

For implementation purposes, we discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area
$\Delta$. The integral expression in Eq. (\ref{measVAR}) and
(\ref{sur}) are then approximated by sums over all grid cells. This
grid is identical to the possible set of design locations and defined
as a waypoint graph \kc{You mean the
  centroids of the cells are part of the graph?}. Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by: 
\begin{equation}\label{prior}
    \bxi = (\xi_a(\bx_1),\xi_b(\bx_1),\ldots,\xi_a(\bx_n),\xi_b(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ holds the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ positive definite matrix $\bSigma$ holds the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

We denote the data at $n_y$ locations by
$\by=(y_{a,1},y_{b,1},\ldots,y_{a,n_y},y_{b,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by: 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bA \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bA\bxi)^T\bR^{-1}(\by-\bA\bxi)},
\end{equation}
where $\bA$ is an $2n_y \times 2n$ matrix holding '$1$'s on the
measurement indices on the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_a$ and $r^2_b$ which are measurement error variances of
temperature and salinity observations.

The GP mean and covariance in Eq. (\ref{prior}) are moderated \kc{What
  does 'moderated' mean?} when more information is available. The
updated distribution for temperature and salinity variables on the
spatial grid, given survey data $\by$, is Gaussian with mean and
covariance: 
\begin{eqnarray}\label{gp_upd}
  \bm &=& \bm(\by) = \bmu+\bSigma \bA^t (\bA \Sigma \bA^t+\bR)^{-1}(\by-\bA \bmu),  \\
  \bS &=& \bSigma - \bSigma \bA^t (\bA \bSigma \bA^t+\bR)^{-1} \bA
          \bSigma.\nonumber
\end{eqnarray}

The updated bivariate distribution at a grid location $\bx \in
\mathcal{M}_g$ is then: 
\begin{equation}\label{gp_hat}
\begin{bmatrix}
\xi_a(\bx) \\
\xi_b(\bx)
\end{bmatrix}
 |\by
 \sim N \left( 
\begin{bmatrix} m_{a}(\bx)\\
m_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
s_{a}^2(\bx) & s_{a,b}(\bx)  \\
s_{a,b}(\bx)  & s_{b}^2(\bx)  
\end{bmatrix}
\right),
\end{equation}
with entries extracted from the conditional mean and covariance
expressions in Eq. (\ref{gp_upd}).

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bA \bmu,\bA \bSigma \bA^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before knowing the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bA^t (\bA \bSigma \bA^t+\bR)^{-1} \bA \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{a}(\bx) \\
      \mu_{b}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{a}(\bx) & \psi_{a,b}(\bx)\\
      \psi_{a,b}(\bx) & \psi^2_{b}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed form solution of the design criteria.

@@

\begin{eqnarray}\label{eq:prob_mv0}
 p(\bx) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b) \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_a\\
t_b
\end{bmatrix};
\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{a}^2(\bx) & \sigma_{a}(\bx)\sigma_{b}(\bx) \gamma(\bx)  \\
\sigma_{a}(\bx)\sigma_{b}(\bx) \gamma(\bx)  & \sigma_{b}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
where $\Phi_2$ is the bivariate Gaussian cumulative distribution
function, and the mean and covariance entries are as defined in Eq. (\ref{gp_bar}).

When data $\by$ are available, Eq. (\ref{eq:post_ep}) similarly gives
us:
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by)
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_a\\
t_b
\end{bmatrix};
\begin{bmatrix} m_{a}(\bx)\\
m_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
s_{a}^2(\bx) & s_{a,b}(\bx)  \\
s_{a,b}(\bx)  & s_{b}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters as defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}).

@@

We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity.
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.eps}
  \caption{Density contour plots having different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. x-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean, $\mu_a=t_a=5^o C$ and
salinity $\mu_b=t_b=30$ mg/l. The displayed densities have unit
standard deviations for both temperature and salinity, but we also
study the effect of doubling the standard deviations.

@@

These mean and covariance parameters values were then used in a field
trial, where we explored the algorithm's ability to characterize the
front between the river and fjord water masses.

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task for
  modern marine management and science faced with limited resources to
  understand highly dynamic processes in the water-column. The
  combination of statistics and autonomous
  robotics provides opportunities for studying aspects related to
  experimental design for effective characterization of oceanographic
  phenomena. In this work we develop methods for efficient spatial
  sampling applied to the mapping of coastal processes by
  providing informative descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random
  fields, and derive analytical expressions for the expected Bernoulli
  variance reduction in the excursion set of spatial temperature and
  salinity variables. We show how these can be applicable in static
  designs and sequential designs, building on myopic and look-ahead
  sampling strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches. Results from field
  deployments with an autonomous underwater vehicle are also presented,
  for a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

\begin{itemize}
\item Decisions on where to sample happen sequentially, i.e.  how
  recently acquired information can be assimilated and used to improve
  statistical models and reduce uncertainty associated with where to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there lies a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}, this is
  usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the optimal design) and computability (arriving at a
  solution in reasonable time), which has to be considered in
  practice.
\end{itemize}{}

@@

In this work, we build on joint excursion sets (ES) and excursion probabilities (EP) for spatial design of experiments. The goal is to
design sampling strategies for reducing the spatially integrated
Bernoulli variance (IBV) in the ES indicators.  The main contributions
involve: i) extending ES-based uncertainty criteria to multivariate
GPs, allowing the evaluation to include multiple environmental
parameters such as temperature and salinity, ii) developing methods
for informative environmental sensing and autonomous data collection,
and iii) demonstrating the approach in a real world scenario of
sampling a river plume using an AUV.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a range of scientific payloads to survey the
water column. Typically, they operate at $\sim 1-3$ m/s, can reach
depths from 100--6000 m with an in-water operation time capacity depending on
survey speed, payload sensors and mission design. By deploying AUVs
together with other resources one can augment information obtained
from either ocean models, remote-sensing data, or fixed-location
sensors on buoys. Traditional operations and surveys with AUVs are
usually limited to observations along fixed transects on a spatial
grid, pre-programmed by the human operator. By instead using on-board
algorithms to continuously evaluate, update, and refine future
sampling locations, one can make information gathering adaptive and
contextual to the survey requirements
\citep{das11b,fossuminformation,fossum18b}. The space of sampling
opportunities is still limited to the spatial grid, but now there is
much more flexibility because of adaptation. For instance, an AUV
could reconstruct or modify a survey line based on what temperatures
it is measuring by using embedded methods in decision-theoretic
planning \citep{py10,Rajan12,Rajan12b}.

@@

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater (from the river) and warmer saline waters of the fjord. Excursion sets (ES) and excursion probabilities (EP) are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

We consider a lateral domain, near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes;
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. The generalization to more than two variables is given in Section \ref{gen_k}.

We let $t_T$ be a threshold for temperature and $t_S$ a threshold for salinity, the joint ES is then defined by:
\begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_T(\bx) \leq t_T,\xi_S(\bx) \leq t_S\}.
\end{equation}
The excursions could also be defined as larger than a threshold, or
within a boundary of limits for the two variables; calculations and
interpretations will be similar in these cases.

The ES holds random indicator variables, while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is however
challenging if the probability is close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

If prior knowledge provides a strong indication of cold freshwater
(for instance) in a part of the domain, it is easy to classify, and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification skill. We denote temperature measurements by
$y_{T}(\bx_i)$ and salinity measurements by $y_{S}(\bx_i)$, at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements, denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by). 
\end{equation}

Our goal is to construct adaptive sampling strategies based on continuous evaluation of the EPs. As the EPs provide a description of the boundary uncertainty, the strategies can effectively use this description to prioritize sampling efforts to locations that are ambiguous; making exploration more effective. As the AUV gathers data throughout its mission, it is natural to evaluate the expected reduction in the IBV, conditioned on the survey data. In this way, we can
take the expectation with respect to the planned measurements $\by$;
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey. A criterion for the selection is:
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}), and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria for situations where the underlying model is based on multivariate GPs.

@@

Calculations of the EPs and the IBV require a model specification. We use Gaussian modeling, which provides a practical and popular probabilistic approach to modeling environmental data, allowing a closed form
evaluation of the IBVs, as derived in Section \ref{sec:sur}.

@@

The Gaussian model is used to represent the bivariate random fields of temperature and salinity. At location $\bx \in \mathcal{M}$, the prior bivariate distribution is:
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (Normal) distributed variables
specified by the mean vector and covariance matrix. For the application
of tracking river plumes, the model parameters are specified from a short preliminary survey. The mean varies with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and temperature are assumed to be constant for all locations and within each water mass.

For the spatial dependence among variables, we assume a separable correlation function with the same decay for salinity and temperature,
which is not unrealistic for our case with a river plume as the decay is related to the common water masses (Sect \ref{sec:case_study}). We have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$
otherwise. Moreover, we assume stationary isotropic random fields a priori where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via a Euclidean distance measure $h=\sqrt{||\bx-\bx'||^2}$. With
data and prior knowledge one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however that is left for
future work.

We discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The
grid is used for the waypoint graph setting the possible design locations. It is also used when solving integrals expressions in Eq. (\ref{measVAR}) and
(\ref{sur}) which are then approximated by sums over all grid cells.  Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by: 
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ holds the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ holds the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by: 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is an $2n_y \times 2n$ matrix holding '$1$'s on the
measurement indices on the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

The GP mean and covariance in Eq. (\ref{prior}) are updated when more information is available. The
updated distribution for temperature and salinity variables on the
spatial grid, given survey data $\by$, is Gaussian with mean and
covariance: 
\begin{eqnarray}\label{gp_upd}
  \bm &=& \bm(\by) = \bmu+\bSigma \bG^t (\bG \Sigma \bG^t+\bR)^{-1}(\by-\bG \bmu),  \\
  \bA &=& \bSigma - \bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG
          \bSigma.\nonumber
\end{eqnarray}

The updated bivariate distribution at a grid location $\bx \in
\mathcal{M}_g$ is then: 
\begin{equation}\label{gp_hat}
\begin{bmatrix}
\xi_T(\bx) \\
\xi_S(\bx)
\end{bmatrix}
 |\by
 \sim N \left( 
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}
\right),
\end{equation}
with entries $m_{\cdot}$ and $a_{\cdot}$ extracted from the conditional mean and covariance
expressions in Eq. (\ref{gp_upd}).

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before knowing the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed form solution of the design criteria.

@@

\begin{eqnarray}\label{eq:prob_mv0}
 p(\bx) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S) \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx)\sigma_{S}(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx)\sigma_{S}(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
where $\Phi_2$ is the bivariate Gaussian cumulative distribution
function, and the mean and covariance entries are as defined in Eq. (\ref{gp_bar}).

When data $\by$ are available, Eq. (\ref{eq:post_ep}) similarly gives
us:
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by)
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters as defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}). Here, the probabilities would not only depend on the data but also on the data design defined via matrix $\bG$ in Eq. (\ref{likelihood}). We next derive results that ease comparison of several designs.

@@

We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity.
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.eps}
  \caption{Density contour plots having different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. x-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean, $\mu_T=t_T=5^o C$ and
salinity $\mu_S=t_S=30$ mg/l. The displayed densities have unit
standard deviations for both temperature and salinity, but we also
study the effect of doubling the standard deviations.

@@
